From the function-sheaf dictionary to quasicharacters of $p$-adic tori

### 2015-07-18 dictionary.tex
Catching up

Let $K$ be a local field with finite residue field $\Fq$;
 we continue to assume that $\ell$ is invertible in $\Fq$.
In this section we consider connected commutative algebraic groups over $K$ that admit a N\'eron model, by which we mean a locally finite type N\'eron model.
By \cite{bosch-lutkebohmert-reynaud:NeronModels}*{\S 10.2, Thm 2}, these are precisely the connected commutative algebraic groups over $K$ that contain no subgroup isomorphic to $\mathbb{G}_\text{a}$.
Write $\mathcal{N}_K$ for the full subcategory of the category of algebraic groups consisting of such objects.  This category is additive, and includes all algebraic tori over $K$, abelian varieties over $K$ and unipotent $K$-wound groups.

@@

\begin{theorem}\label{thm:CSXK}
Let $K$ be a local field with residue field $\Fq$, in which $\ell$ is invertible; 
let $R$ be the ring of integers of $K$.
\begin{enumerate}
\labitem{(0)}{x0}
The trace of Frobenius provides a natural transformation between the additive functors
\[
F_{\QCS}: X_K \mapsto \QCSiso{X}
\qquad\text{and}\qquad
F_{\text{}}: X_K \mapsto \Hom_{\text{}}(X(K),\EEx)
\]
as functors from $\mathcal{N}_K$ to the category of commutative groups.
\end{enumerate} 
Regarding this natural transformation, for every $X_K \in \mathcal{N}_K$ with N\'eron model $X$:
\begin{enumerate}[resume]
\labitem{(1)}{x1} there is a canonical short exact sequence of commutative groups 
\[
0 \to \Hh^2(\pi_0(X)_{\bFq},\EEx)^{\Weil{}} \to \QCSiso{X} \to \Hom_\text{}(X(K),\EEx) \to 0;
\] 
\labitem{(2)}{x2} for all quasicharacter sheaves $\cs{F}$, $\cs{F}'$ on $\Gr_R(X)$, and for every $\rho \in \Hom(\cs{F},\cs{F}')$, either $\rho$ is trivial or $\rho$ is an isomorphism;
\labitem{(3)}{x3} for all quasicharacter sheaves $\cs{F}$ for $X_K$, there is a canonical isomorphism
\[
\Aut(\cs{F}) \iso \Hom((\pi_0(X)_{\bFq})_{\Weil{}},\EEx).
\]
\end{enumerate}
\end{theorem}

@@

\begin{corollary}\label{cor:CS_tori}
Let $T_K$ be an algebraic toris over $K$ and let $T$ be a Neron model for $T_K$.
The following is a commutative diagram of exact sequences.
\[
  \begin{tikzcd}[row sep=20, column sep=18]
{}  & 0 \arrow{d} & 0 \arrow{d} &  & \\ 
   & \Hh^2(X_*(T_K)_{\mathcal{I}_K},\EEx)^{\Weil{}}  \arrow{d} & \arrow{d} \Hh^2(X_*(T_K)_{\mathcal{I}_K},\EEx)^{\Weil{}} & 0 \arrow{d} & \\
    0 \rar & \QCSiso{\pi_0(T)} \arrow{r}{\pi_0^*} \dar{\TrFrob{\pi_0(T)}}
    & \QCSiso{T} \arrow{r}{\iota_0^*}  \arrow{d}{\TrFrob{\Gr_R(T)}} & \QCSiso{T^0} \arrow{r} \arrow{d}{\TrFrob{\Gr_R(T)^0}} & 0\\
    0 \arrow{r} & \Hom(\pi_0(T)(\Fq), \EEx) \arrow{r}{\text{infl'n}} \arrow{d}{\iso}
    & \arrow{d}{\iso} \Hom_\text{}(T(K),\EEx) \arrow{r}{\text{rest'n}} & \Hom_\text{}(T^0(R),\EEx) \arrow{r} \arrow{d}{\iso} & 0\\
 &  0  & 0 & 0 & 
  \end{tikzcd}
 \]
 \end{corollary}
\begin{proof} 
The horizontal sequence of groups coming from categories of quasicharacter sheaves is exact by Proposition~\ref{prop:middleexact}, together with the observation that the functors $\pi_0^*$ and $\iota^*$ perverse limits.
It is elementary that the horizontal sequence of quasicharacters is exact.
The diagram commutes by Lemma~\ref{lem:pullback}.
Accordingly, by Theorem~\ref{thm:CSXK}, the kernel of $\TrFrob{T}$ is $\Hh^2(\pi_0(T)_{\bFq},\EEx)^{\Weil{}}$. 
By \cite{bitan:discriminant}*{Eq 3.1}, the special fibre of the component group scheme for $T$ is given by
\[
 \pi_0(T)_{\bFq} = X_*(T_K)_{\mathcal{I}_K},
\]
where $X_*(T_K)$ is the cocharacter lattice of $T_K$ and $\mathcal{I}_K$ is the inertia group for $K$.
Thus,
\[
\Hh^2(\pi_0(T)_{\bFq},\EEx)^{\Weil{}} = \Hh^2(X_*(T_K)_{\mathcal{I}_K},\EEx)^{\Weil{}}.
\] 
which coincides with the kernel of $\TrFrob{\pi_0(T)}$ since $\pi_0(\pi_0(T))_{\bFq} = \pi_0(T)_{\bFq}$.
\end{proof}

@@

\begin{lemma}\label{lem:LLCT}
The following diagram commutes. 
\[
\begin{tikzcd}
   0 \arrow{r} & \Hom(\pi_0(T)(\Fq), \EEx) \arrow{r}{\text{infl'n}} \arrow{d}{\iso}
    & \arrow{d}{\iso} \Hom_\text{}(T(K),\EEx) \arrow{r}{\text{rest'n}} & \Hom_\text{}(T^0(R),\EEx) \arrow{r} \arrow{d}{\iso} & 0\\    
    0 \arrow{r}  
 & \Hh^1(\Weil{}, \check{T}_\ell^{\mathcal{I}_K}) \arrow{r}{\text{infl'n}}
 & \Hh^1(\Weil{K}, \check{T}_\ell) \arrow{r}{\text{rest'n}} 
 & \Hh^1(\mathcal{I}_K, \check{T}_\ell)^{\Weil{}} \arrow{r}  
 & 0
\end{tikzcd}
\] 
\end{lemma}
\begin{proof}
Since $\pi_0(T)_{\bFq} = X_*(T_K)_{\mathcal{I}_K}$ by \cite{bitan:discriminant}*{Eq 3.1}, the isomorphism on the left can be seen directly:
\[\Hom(\pi_0(T)(\Fq), \EEx)
= \Hom((X_*(T)_{\mathcal{I}_K})^{\Weil{}}, \EEx)
= \Hom(X_*(T)_{\mathcal{I}_K}, \EEx)_{\Weil{}} 
%= \Hh^1(\Weil{}, \Hom(X_*(T), \EEx)^{\mathcal{I}_K}) 
= \Hh^1(\Weil{}, \check{T}_\ell^{\mathcal{I}_K}).
\]
The middle isomorphism is the Langlands correspondence for $p$-adic tori, itself a fairly direct consequence of local class field theory; see \cite{yu:09a}, for example. The isomorphism 
\[
\Hom_\text{}(T^0(R),\EEx) \iso \Hh^1(\mathcal{I}_K, \check{T}_\ell)^{\Weil{}}
\] 
and the commutativity of the diagram is then a consequence of the inflation-restriction exact sequence and the fact that $\Weil{}$ has cohomological dimension $1$.
\end{proof}

Combining Corollary~\ref{cor:CS_tori} with Lemma~\ref{lem:LLCT} now produces the following commutative diagram of exact sequences:
\begin{equation}\label{eq:GLCT}
  \begin{tikzcd}[row sep=20, column sep=15]
{}  & 0 \arrow{d} & 0 \arrow{d} &  & \\ 
   & \Hh^0(\Weil{}, \Hh^2(X_*(T_K)_{\mathcal{I}_K},\EEx))  \arrow{d} & \arrow{d} \Hh^0(\Weil{}, \Hh^2(X_*(T_K)_{\mathcal{I}_K},\EEx)) & 0 \arrow{d} & \\
    0 \rar & \QCSiso{\pi_0(T)} \arrow{r}{\pi_0^*} \arrow{d}
    & \QCSiso{T} \arrow{r}{\iota_0^*}  \arrow{d} & \QCSiso{T^0} \arrow{r} \arrow{d} & 0\\
   0 \arrow{r}  
 & \Hh^1(\Weil{}, \Hh^0(\mathcal{I}_K, \check{T}_\ell)) \arrow{r}{\text{infl'n}} \arrow{d}
 & \Hh^1(\Weil{K}, \check{T}_\ell) \arrow{r}{\text{rest'n}} \arrow{d}
 & \Hh^0(\Weil{}, \Hh^1(\mathcal{I}_K, \check{T}_\ell)) \arrow{r} \arrow{d}  
 & 0\\
 &  0  & 0 & 0 & 
  \end{tikzcd}
 \end{equation}
It is natural to ask if the vertical surjections can be defined directly, without making use of local class field theory.  
Even the case $T_K = \Gm{K}$ is remarkable, in which case \eqref{eq:GLCT} becomes

In fact, the answer is yes; as such, these maps provide a geometric perspective on the Langlands correspondence for $p$-adic tori.

As in \cite{yu:09a}, the general case can be built up from the case $T_K = \Gm{K}$.

or by a small part of Proposition~\ref{prop:etale-iso}

[insert discussion of Takashi Suzuki's work]

>>>>

Let $K$ be a local field with finite residue field $\Fq$; we will now denote the group $\Weil{}$ by $\Weil{\Fq}$.
We continue to assume that $\ell$ is invertible in $\Fq$. 
In this section we consider connected commutative algebraic groups over $K$ that admit a N\'eron model, by which we mean a locally finite type N\'eron model.
By \cite{bosch-lutkebohmert-reynaud:NeronModels}*{\S 10.2, Thm 2}, these are precisely the connected commutative algebraic groups over $K$ that contain no subgroup isomorphic to $\mathbb{G}_\text{a}$.
Write $\mathcal{N}_K$ for the full subcategory of the category of algebraic groups consisting of such objects.  This category is additive, and includes all algebraic tori over $K$, abelian varieties over $K$ and unipotent $K$-wound groups.

@@

\begin{theorem}\label{thm:CSXK}
Let $K$ be a local field with residue field $\Fq$, in which $\ell$ is invertible; 
let $R$ be the ring of integers of $K$.
\begin{enumerate}
\labitem{(0)}{x0}
The trace of Frobenius provides a natural transformation between the additive functors
\[
F_{\QCS}: X_K \mapsto \QCSiso{X}
\qquad\text{and}\qquad
F_{\text{}}: X_K \mapsto \Hom_{\text{}}(X(K),\EEx)
\]
as functors from $\mathcal{N}_K$ to the category of commutative groups.
\end{enumerate} 
Regarding this natural transformation, for every $X_K \in \mathcal{N}_K$ with N\'eron model $X$:
\begin{enumerate}[resume]
\labitem{(1)}{x1} there is a canonical short exact sequence of commutative groups 
\[
0 \to \Hh^2(\pi_0(X)_{\bFq},\EEx)^{\Weil{\Fq}} \to \QCSiso{X} \to \Hom_\text{}(X(K),\EEx) \to 0;
\] 
\labitem{(2)}{x2} for all quasicharacter sheaves $\cs{F}$, $\cs{F}'$ on $\Gr_R(X)$, and for every $\rho \in \Hom(\cs{F},\cs{F}')$, either $\rho$ is trivial or $\rho$ is an isomorphism;
\labitem{(3)}{x3} for all quasicharacter sheaves $\cs{F}$ for $X_K$, there is a canonical isomorphism
\[
\Aut(\cs{F}) \iso \Hom((\pi_0(X)_{\bFq})_{\Weil{\Fq}},\EEx).
\]
\end{enumerate}
\end{theorem}

@@

\begin{corollary}\label{cor:CS_tori}
Let $T_K$ be an algebraic toris over $K$ and let $T$ be a Neron model for $T_K$.
The following is a commutative diagram of exact sequences.
\[
  \begin{tikzcd}[row sep=20, column sep=18]
{}  & 0 \arrow{d} & 0 \arrow{d} &  & \\ 
   & \Hh^2(X_*(T_K)_{\mathcal{I}_K},\EEx)^{\Weil{\Fq}}  \arrow{d} & \arrow{d} \Hh^2(X_*(T_K)_{\mathcal{I}_K},\EEx)^{\Weil{\Fq}} & 0 \arrow{d} & \\
    0 \rar & \QCSiso{\pi_0(T)} \arrow{r}{\pi_0^*} \dar{\TrFrob{\pi_0(T)}}
    & \QCSiso{T} \arrow{r}{\iota_0^*}  \arrow{d}{\TrFrob{\Gr_R(T)}} & \QCSiso{T^0} \arrow{r} \arrow{d}{\TrFrob{\Gr_R(T)^0}} & 0\\
    0 \arrow{r} & \Hom(\pi_0(T)(\Fq), \EEx) \arrow{r}{\text{infl'n}} \arrow{d}{\iso}
    & \arrow{d}{\iso} \Hom_\text{}(T(K),\EEx) \arrow{r}{\text{rest'n}} & \Hom_\text{}(T^0(R),\EEx) \arrow{r} \arrow{d}{\iso} & 0\\
 &  0  & 0 & 0 & 
  \end{tikzcd}
 \]
 \end{corollary}
\begin{proof} 
The horizontal sequence of groups coming from categories of quasicharacter sheaves is exact by Proposition~\ref{prop:middleexact}, together with the observation that the functors $\pi_0^*$ and $\iota^*$ perverse limits.
It is elementary that the horizontal sequence of quasicharacters is exact.
The diagram commutes by Lemma~\ref{lem:pullback}.
Accordingly, by Theorem~\ref{thm:CSXK}, the kernel of $\TrFrob{T}$ is $\Hh^2(\pi_0(T)_{\bFq},\EEx)^{\Weil{}}$. 
By \cite{bitan:discriminant}*{Eq 3.1}, the special fibre of the component group scheme for $T$ is given by
\[
 \pi_0(T)_{\bFq} = X_*(T_K)_{\mathcal{I}_K},
\]
where $X_*(T_K)$ is the cocharacter lattice of $T_K$ and $\mathcal{I}_K$ is the inertia group for $K$.
Thus,
\[
\Hh^2(\pi_0(T)_{\bFq},\EEx)^{\Weil{\Fq}} = \Hh^2(X_*(T_K)_{\mathcal{I}_K},\EEx)^{\Weil{\Fq}}.
\] 
which coincides with the kernel of $\TrFrob{\pi_0(T)}$ since $\pi_0(\pi_0(T))_{\bFq} = \pi_0(T)_{\bFq}$.
\end{proof}

@@

\begin{lemma}\label{lem:LLCT}
The following diagram commutes. 
\[
\begin{tikzcd}
   0 \arrow{r} & \Hom(\pi_0(T)(\Fq), \EEx) \arrow{r}{\text{infl'n}} \arrow{d}{\iso}
    & \arrow{d}{\iso} \Hom_\text{}(T(K),\EEx) \arrow{r}{\text{rest'n}} & \Hom_\text{}(T^0(R),\EEx) \arrow{r} \arrow{d}{\iso} & 0\\    
    0 \arrow{r}  
 & \Hh^1(\Weil{\Fq}, \check{T}_\ell^{\mathcal{I}_K}) \arrow{r}{\text{infl'n}}
 & \Hh^1(\Weil{K}, \check{T}_\ell) \arrow{r}{\text{rest'n}} 
 & \Hh^1(\mathcal{I}_K, \check{T}_\ell)^{\Weil{\Fq}} \arrow{r}  
 & 0
\end{tikzcd}
\] 
\end{lemma}
\begin{proof}
Since $\pi_0(T)_{\bFq} = X_*(T_K)_{\mathcal{I}_K}$ by \cite{bitan:discriminant}*{Eq 3.1}, the isomorphism on the left can be seen directly:
\[\Hom(\pi_0(T)(\Fq), \EEx)
= \Hom((X_*(T)_{\mathcal{I}_K})^{\Weil{\Fq}}, \EEx)
= \Hom(X_*(T)_{\mathcal{I}_K}, \EEx)_{\Weil{\Fq}} 
%= \Hh^1(\Weil{}, \Hom(X_*(T), \EEx)^{\mathcal{I}_K}) 
= \Hh^1(\Weil{\Fq}, \check{T}_\ell^{\mathcal{I}_K}).
\]
The middle isomorphism is the Langlands correspondence for $p$-adic tori, itself a fairly direct consequence of local class field theory; see \cite{yu:09a}, for example. The isomorphism 
\[
\Hom_\text{}(T^0(R),\EEx) \iso \Hh^1(\mathcal{I}_K, \check{T}_\ell)^{\Weil{\Fq}}
\] 
and the commutativity of the diagram is then a consequence of the inflation-restriction exact sequence and the fact that $\Weil{}$ has cohomological dimension $1$.
\end{proof}

Combining Corollary~\ref{cor:CS_tori} with Lemma~\ref{lem:LLCT} now produces the following commutative diagram of exact sequences:
\begin{equation}\label{eq:GLCT}
  \begin{tikzcd}[row sep=20, column sep=15]
{}  & 0 \arrow{d} & 0 \arrow{d} &  & \\ 
   & \Hh^2(X_*(T_K)_{\mathcal{I}_K},\EEx)^{\Weil{\Fq}}  \arrow{d} & \arrow{d} \Hh^2(X_*(T_K)_{\mathcal{I}_K},\EEx)^{\Weil{\Fq}} & 0 \arrow{d} & \\
    0 \rar & \QCSiso{\pi_0(T)} \arrow{r}{\pi_0^*} \arrow{d}
    & \QCSiso{T} \arrow{r}{\iota_0^*}  \arrow{d} & \QCSiso{T^0} \arrow{r} \arrow{d} & 0\\
   0 \arrow{r}  
 & \Hh^1(\Weil{\Fq}, \check{T}_\ell^{\mathcal{I}_K}) \arrow{r}{\text{infl'n}} \arrow{d}
 & \Hh^1(\Weil{K}, \check{T}_\ell) \arrow{r}{\text{rest'n}} \arrow{d}
 & \Hh^1(\mathcal{I}_K, \check{T}_\ell)^{\Weil{\Fq}} \arrow{r} \arrow{d}  
 & 0\\
 &  0  & 0 & 0 & 
  \end{tikzcd}
 \end{equation}
It is natural to ask if the vertical surjections can be defined directly, without making use of local class field theory.  
The case $T_K = \Gm{K}$ is already very interesting, in which case \eqref{eq:GLCT} becomes
\begin{equation}\label{eq:Gm}
\begin{tikzcd}
    0 \rar & \QCSiso{\ZZ} \arrow{r}{\pi_0^*} \arrow{d}{\iso}
    & \QCSiso{T} \arrow{r}{\iota_0^*}  \arrow{d}{\iso} & \QCSiso{\Gm{R}} \arrow{r} \arrow{d}{\iso} & 0\\
    0 \arrow{r}  
 & \Hom(\Weil{\Fq}, \EEx) \arrow{r}{\text{infl'n}}
 & \Hom(\Weil{K}, \EEx) \arrow{r}{\text{rest'n}} 
 & \Hom(\mathcal{I}_K, \EEx) \arrow{r}  
 & 0
\end{tikzcd}
\end{equation}
We have seen that the isomorphism $\QCS(\ZZ)\iso \Hom(\Weil{\Fq}, \EEx)$ is elementary. 
The isomorphism $\QCSiso{\Gm{R}}\iso \Hom(\mathcal{I}_K, \EEx)$ may be deduced directly from the class field theory of Serre and Hezewinkel. The monoidal category of character sheaves on $\Gr^R_n(\Gm{R})$ is canonically equivalent to the category of $\ell$-adic characters of the equivariant etale fundamental group $\pi_1^\text{eq}(\Gr^R_n(\Gm{R}),\bFq)$.
%and, using the Lang isogeny, that there is a canonical isomorphism 
%\[\pi_1^\text{eq}(\Gr^R_n(\Gm{R}),\bFq) \iso \Gr^R_n(\Gm{R})(\Fq) = R_n^*.\]
Accordingly, quasicharacter sheaves on $\Gm{R}$ may be apprehended as $\ell$-adic characters of
the inverse limit of the groups $\pi_1^\text{eq}(\Gr^R_n(\Gm{R}),\bFq)$, as $n$ ranges over $\NN$.
By \cite{serre:isogenies}, there is a canonical isomorphism 
\[
\varprojlim_{n\in \NN} \pi_1^\text{eq}(\Gr^R_n(\Gm{R}),\bFq) \iso \mathcal{I}_K^\text{ab};
\]
this isomorphism is {\it not} restricted to the case when $K$ and $\Fq$ have equal characteristic.
This last isomorphism now provides the isomorphism $\QCSiso{\Gm{R}}\iso \Hom(\mathcal{I}_K, \EEx)$ in \eqref{eq:Gm}. 
Since the sequences are split, this gives \eqref{eq:Gm}, geometrically. 
We have been glib here because all this is treated carefully and beautifully in \cite{suzuki-yoshida:12a}. 
Indeed, in that paper, an isomorphism is constructed directly between the equivariant fundamental group of $\Gr_R(T)$ (with respect to a site defined in \cite{suzuki-yoshida:12a}) when $T$ is a Neron model of $\Gm{K}$, and $\Weil{K}^\text{ab}$, improving greatly on the arguments of this paragraph.

We expect that the general case of \eqref{eq:GLCT}, where $K$ is any local field and $T_K$ is any torus over $K$, may be deduced from the case \eqref{eq:Gm}, by an adaptation of arguments from \cite{yu:09a}. 
In anticipation of work in that direction, Section~\ref{ssec:wrK} provides a key tool for that argument.

We close this section with some further remarks on sites.
